A thermodynamic view of dusty protoplanetary disks

### 2017-05-02 summary.tex
misc edits from sect 4 to sect 8.3

\subsubsection{Classic gravitational instabilities} %vertical structure 
The addition of dust enhances gravitational
instability (GI). This is because dust particles contribute to the
total disk mass but not thermal pressure, which effectively lowers the
disk temperature \citep[][]{thompson88,shi13}. For typical dust-loading 
$\epsilon, Z\ll1$, this effect is unimportant. However, if $\epsilon$ is
large (e.g. due to dust settling) then the reduced temperature $\tilde{T}$ may 
be lowered to enable instability.

As noted in \S\ref{loc_iso_eos}, dust settling causes 
$\tilde{T}$ to \emph{increase} away from the midplane. This contrasts
to previous studies of GI in vertically stratified disks
 where the temperature decreases
from the midplane \citep[e.g.][]{mamat10, kim12,lin14c}. 
While we expect only the total surface density and
characteristic temperatures are relevant to stability  
\citep{toomre64}, a non-trivial effective vertical temperature
structure, induced by dust, may modify the vertical structure of 3D
waves and unstable modes. We will examine this linear problem in a following 
study.

@@

\subsubsection{Rossby wave instability}
The Rossby wave instability (RWI) is a non-axisymmetric, 2D shear
instability that operates in thin disks when it has radial structure
\citep{lovelace99,li00}. These studies consider adiabatic pure gas and 
show instability is possible if there is an extrema in the generalized
potential vorticity 
\begin{align}
 \mathcal{V} = \frac{\kappa^2}{2\Omega\Sigma}\mathcal{S}^{-2/\gamma},  
\end{align} 
where $\mathcal{S} = \Pi/\Sigma^\gamma$. Here, $\Sigma$ is the total
surface density and $\Pi$ is the vertically integrated pressure.

From the dusty/adiabatic gas equivalence, we deduce the corresponding
condition for RWI in a (locally isothermal or polytroipc) dusty gas is
also given by extrema in $\mathcal{V}$ with $\mathcal{S} = \exp{\left(\Gamma
S_\mathrm{eff}\right)}$ and $\tepsilon = \Sigma_\mathrm{d}/\Sigma$. Since
the effective entropy reflects the dust distribution, we expect that
RWI may also be triggered by extrema in the dust-to-gas ratio, e.g. 
narrow dust rings. This would lead to direction formation of dusty
vortices.

@@

\subsubsection{Convective overstability}
The `convective overstability' (ConO) operates in dust-free, non-adiabatic disks 
when there is an unstable radial buoyancy frequency ($N_r^2 < 0 $)
combined with finite thermal relaxation time, $t_\mathrm{cool}\sim
\OmK^{-1}$. This lead to growing epicycles
\citep{klahr14,lyra14,latter16}.  
%with 
%growth rate $\propto  -N_r^2$ 
% Indeed, the effective radial buoyancy frequency in our dusty disk 
% with negligble background density gradient is unstable, 
% \begin{align}
%   N_r^2 \equiv -\frac{1}{\rho}\frac{\p P}{\p r}\frac{\p S_\mathrm{eff}}{\p r} =
%   -\frac{\rho}{P}F_r^2<0.  
% \end{align}

The conditions required for ConO might be realized in a locally
isothermal dusty disk. Since $\p_rP < 0$ in typical disk models, 
if the dust-induced radial entropy gradient is negative, then
$N_r^2  =  F_r \p_r\seff < 0$. In 
fact, the typical disk models used to study the streaming
instability have zero background radial density gradient, in which
case  $N_r^2= -\rho F_r^2/P$.

Locally isothermal disks have $\tcool=0$, so the pure gas ConO cannot
exist. However, we have shown that dust-gas drag provides an effective
energy source/sink to the mixture. It would be
interesting to explore whether or not dust-gas drag can provide the
required `cooling' to enable a dusty version of the convective
overstability.

@@

The `zombie vortex instability' (ZVI) is a non-axisymmetric,
non-linear (or finite-amplitude) instability in gas disks
\citep{marcus15}. It operates by exciting `critical layers' which role
up into vortices, exciting further critical layers. The result is an
array of vortices. These critical
layers are locations where the intrinsic wave frequency of a
disturbance matches the local vertical buoyancy frequency
\citep{marcus13}.

Since buoyancy is critical for the ZVI, it requires an almost
purely adiabatic gas ($\tcool\OmK\gg 1$). This limits its
applicability to the inner most parts of typical protoplanetary disks
\citep{lesur16}.

However, we have show that dust-loading an isothermal disk also
induces an effective buoyancy. This effect is maximized when dust is
perfectly coupled to the gas. It is thus natural to ask whether or not
the required adiabatic conditions for ZVI can be realized through
dust-loading, and thus produce a `dusty ZVI'.

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas. One need not add any new 
equations to evolve the dust component: 
%In fact, no
%modification is needed  
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
 the standard energy equation substitutes for the dust continuity
equation. When $c_s$ or $K$ is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15}, which is also
featured in many hydrodynamics codes.

\subsection{Higher order corrections}
%{\bf notes for self. maybe exclude in final. }

@@

\section{Summary}\label{summary}
In this paper we develop an analogy between isothermal 
dusty-gas and adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $\tstop \to 0$,  
 dust is entrained in 
the gas. Then the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the flow. This is analogous to entropy being conserved
following a adiabatic, pure gas.

>>>>

\subsubsection{Classic gravitational instabilities} %vertical structure 
The addition of dust enhances gravitational
instability (GI) because dust particles contribute to the
total disk mass but not thermal pressure, which effectively lowers the
disk temperature \citep[][]{thompson88,shi13}. For typical dust-loading 
$\epsilon, Z\ll1$, this effect is unimportant. However, if $\epsilon$ is
large (e.g. due to dust settling) then the reduced temperature
$\widetilde{T}$ may be lowered to enable instability.

As noted in \S\ref{loc_iso_eos}, dust settling causes 
$\widetilde{T}$ to \emph{increase} away from the midplane. This contrasts
to previous studies of GI in vertically stratified disks
 where the temperature decreases
from the midplane \citep[e.g.][]{mamat10, kim12,lin14c}. 
While we expect only the total surface density and
characteristic temperatures are relevant to stability  
\citep{toomre64}, a non-trivial vertical temperature
structure, induced by dust, may modify the vertical structure of 3D
waves and unstable modes. We will examine this problem in a following 
study.

@@

\subsubsection{Rossby wave instability}
The Rossby wave instability (RWI) is a non-axisymmetric, 2D shear
instability that operates in thin disks when it has radial structure
\citep{lovelace99,li00}. These studies consider adiabatic pure gas and 
show instability is possible if there is an extrema in the generalized
potential vorticity 
%\begin{align}
 $\mathcal{V}_\mathrm{g} =
 \kappa^2\mathcal{S}_\mathrm{g}^{-2/\gamma}/2\Omega\Sigma_\mathrm{g}$, 
%\end{align} 
where $\mathcal{S}_\mathrm{g} = P/\Sigma_\mathrm{g}^\gamma$. Here $P$
should be interpreted as the vertically integrated pressure. The
non-linear result of the RWI is vortex formation \citep{li01}.

The dusty/adiabatic gas analogy imply the condition for RWI in a
polytropic dusty gas is given by extrema in  
\begin{align}
  \mathcal{V} =
  \frac{\kappa^2}{2\Omega\Sigma}\frac{1}{K^{2/\Gamma}}\left(1+
    \frac{\Sigma_\mathrm{d}}{\Sigma_\mathrm{g}}\right)^2.    
\end{align}
Thus we expect that RWI may also be triggered by extrema 
in the dust-to-gas ratio, e.g. narrow dust rings/gaps. This may lead
to direct formation of dusty vortices, as opposed to dust-trapping by  
a pre-existing gas vortex \citep{barge95,lyra13}.

@@

\subsubsection{Convective overstability}
The `convective overstability' (ConO) operates in pure gas, non-adiabatic disks 
when there is a radial buoyancy frequency such that $N_r^2\equiv
F_r\p_rS/C_P  < 0 $ and cooling times $t_\mathrm{cool}\sim \OmK^{-1}$. This
combination leads to growing epicycles
\citep{klahr14,lyra14,latter16}.     
%with 
%growth rate $\propto  -N_r^2$ 
% Indeed, the effective radial buoyancy frequency in our dusty disk 
% with negligble background density gradient is unstable, 
% \begin{align}
%   N_r^2 \equiv -\frac{1}{\rho}\frac{\p P}{\p r}\frac{\p S_\mathrm{eff}}{\p r} =
%   -\frac{\rho}{P}F_r^2<0.  
% \end{align}

%The disk structure required for ConO might be realized in a locally
%isothermal dusty disk.

Now consider a strictly isothermal dusty disk where the relevant 
entropy is dust-induced, $\seff =-\ln{(1+\rhod/\rhog)}$. 
%The dusty/adiabatic gas
%analogy implies the replacement $S\to\seff = -\ln{(1+\rhod/\rhog)}$ in
%$N_r$. 
Since $F_r\propto -\p_rP > 0$ in typical disk models, ConO would
require $\p_r\seff<0$, implying the dust-to-gas ratio increases
\emph{outwards}. This might be realized at special locations in
protoplanetary disks.

%fact, the typical disk models used to study the streaming
%instability have zero background radial density gradient, in which
%case  $N_r^2= -\rho F_r^2/P < 0$.

Isothermal disks have $\tcool=0$, so the pure gas ConO cannot
exist. However, we have shown that dust-gas drag provides an effective
energy source/sink to the mixture (\S\ref{energy_analogy}). It would be
interesting to explore whether or not dust-gas drag can provide the 
required `cooling' to enable a dusty version of the convective 
overstability.

@@

The `zombie vortex instability' (ZVI) is a non-axisymmetric,
non-linear instability discovered in pure gas disks
\citep{marcus15,umurhan16d}. It operates by having finite-amplitude perturbations
exciting `critical layers' which role 
up into vortices, exciting further critical layers. The process
repeats itself and the result is an array of vortices. These critical 
layers are locations where the intrinsic wave frequency of a 
disturbance matches the local vertical buoyancy frequency 
\citep{marcus13}.

Since the necessary physical ingredient for the ZVI is vertical
buoyancy, it requires an almost purely adiabatic gas ($\tcool\OmK\gg
1$). This limits its applicability to typical protoplanetary disks at
$\lesssim 1$AU \citep{lesur16, malygin17}. However, this consideration
assumes a dust-free disk.

On the other hand, we have show that dust-loading an isothermal disk
also induces an effective buoyancy. This effect is maximized when dust is
perfectly coupled to the gas. It is thus natural to ask whether or not
the required adiabatic conditions for ZVI can be realized through
dust-loading, and thus produce a `dusty ZVI'. This may allow the ZVI
to operate at $\gtrsim 1$AU.

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas. One need not add any new 
equations to evolve the dust component: 
%In fact, no
%modification is needed  
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
 the standard energy equation substitutes for the dust continuity
equation. When $c_s$ or $K$ is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15} or thermal
conduction, which is also common in modern hydrodynamics codes.

We have taken advantage of the dusty/adiabatic gas equivalence to 
convert the popular
\textsc{Pluto}\footnote{\url{http://plutocode.ph.unito.it/}}      
hydrodynamics code into a dusty gas dynamics code appropriate for
simulating protoplanetary disks coupled to small dust. We will apply 
our modified code --- d\textsc{Pluto} --- to the streaming instability
and disk-planet interactions in an upcoming publication (Lin et al.,
in preparation).

\subsection{Higher order corrections}
{\bf notes for self. maybe exclude in final. }

@@

\section{Summary}\label{summary}
In this paper we develop a physical analogy between isothermal 
dusty-gas and adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $\tstop \to 0$,  
 dust is entrained in 
the gas. Then the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the flow. This property is analogous to entropy being conserved
following a adiabatic, pure gas.
